cardinalR: Generating interesting high-dimensional data structures

### 2025-04-26 paper-cardinalR.Rmd
Added data generation

We illustrate the use of the package by generating five clusters of different shapes, positioned with equal inter-cluster distances in $4\text{-}D$. Using this synthetic dataset, we will evaluate the performance of six popular dimension reduction techniques: Principal Component Analysis (PCA) [@jolliffe2011], t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021].

```{r gen-five-clust-data}

>>>>

This section illustrates the use of package by generating a synthetic dataset to evaluate the performance of six popular dimension reduction techniques: Principal Component Analysis (PCA) [@jolliffe2011], t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021].

We generated a dataset of five clusters, positioned with equal inter-cluster distances in $4\text{-}D$ space. The five clusters have different geometric structures and each contain different number of points. Specifically, the helical spiral cluster includes $2250$ points and was generated with a scale parameter of $0.4$. The hemisphere cluster consists of $1500$ points with a scale parameter of $0.35$. The uniform cube-shaped cluster contains $750$ points and uses a scale parameter of $0.3$. The blunted cone cluster includes $1250$ points, generated with a scale parameter of $1$. Finally, the Gaussian-shaped cluster contains $1750$ points and was generated with a scale parameter of $0.3$.

```{r gen-five-clust-data, echo=TRUE}

### 2025-04-26 paper-cardinalR.tex
Added data generation

The paper is organized as follows. In next section, introduces the implementation of \texttt{cardinalR} package on CRAN and GitHub, including demonstration of the package's key functions. We illustrate how a clustering data structure affect the Non-linear dimension reductions in \textbf{Example} section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for use of our data collection.

@@

\caption{\label{tab:datasets-tb-pdf}cardinalR data sets}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
data & explanation\\
\midrule
mobius\_clust\_data & \\
mobius\_clust\_tsne\_param1 & \\
mobius\_clust\_tsne\_param2 & \\
mobius\_clust\_tsne\_param3 & \\
mobius\_clust\_umap\_param1 & \\
\addlinespace
mobius\_clust\_umap\_param2 & \\
mobius\_clust\_umap\_param3 & \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{verbatim}
gen_multicluster(
  n = c(200, 300, 500), p = 4, k = 3,
  loc = matrix(c(
    0, 0, 0, 0,
    5, 9, 0, 0,
    3, 4, 10, 7
  ), nrow = 4, byrow = TRUE),
  dim_weights = dim4_weights,
  scale = c(3, 1, 2),
  shape = c("gaussian", "bluntedcorn", "unifcube"),
  rotation = rotations_4d,
  is_bkg = FALSE
)
\end{verbatim}

@@

\caption{\label{tab:add-tb-pdf}cardinalR additional functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_noisedims & \\
gen\_bkgnoise & \\
randomize\_rows & \\
gen\_nproduct & \\
gen\_nsum & \\
\addlinespace
gen\_wavydims1 & \\
gen\_wavydims2 & \\
gen\_wavydims3 & \\
gen\_rotation & \\
gen\_globalminmax & \\
\addlinespace
scale\_data & \\
\bottomrule
\end{tabular}
\end{table}

\section{Example}\label{example}

Add one or two datasets and evaluate how it will be useful\ldots. (one with NLDR, one with clustering)

>>>>

The paper is organized as follows. In next section, introduces the implementation of \texttt{cardinalR} package on CRAN and GitHub, including demonstration of the package's key functions. We illustrate how a clustering data structure affect the dimension reductions in \textbf{Application} section. Finally, we give a brief conclusion of the paper and discuss potential opportunities for use of our data collection.

@@

\caption{\label{tab:datasets-tb-pdf}cardinalR data sets}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
data & explanation\\
\midrule
mobiusgau & Simulated data with a Mobius and a Gaussian in 4-D space.\\
mobiusgau\_tsne1 & The tSNE embedding with perplexity \$15\$ for mobiusgau.\\
mobiusgau\_tsne2 & The tSNE embedding with perplexity \$30\$ for mobiusgau.\\
mobiusgau\_tsne3 & The tSNE embedding with perplexity \$5\$ for mobiusgau.\\
mobiusgau\_umap1 & The UMAP embedding with number of neighbors \$15\$ and minimum distance \$0.1\$ for mobiusgau.\\
\addlinespace
mobiusgau\_umap2 & The UMAP embedding with number of neighbors \$30\$ and minimum distance \$0.08\$ for mobiusgau.\\
mobiusgau\_umap3 & The UMAP embedding with number of neighbors \$5\$ and minimum distance \$0.9\$ for mobiusgau.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:main-tb-pdf}The main arguments for `gen_multicluster()`.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Argument & Explanation\\
\midrule
n & A numeric vector representing the sample sizes.\\
p & A numeric value representing the number of dimensions.\\
k & A numeric value representing the number of clusters.\\
loc & A numeric matrix representing the locations/centroids of clusters.\\
scale & A numeric vector representing the scaling factors of clusters.\\
\addlinespace
shape & A character vector representing the shapes of clusters.\\
rotation & A numeric list which contains plane and the corresponding angle along that plane for each cluster.\\
is\_bkg & A Boolean value representing the background noise should exist or not.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:add-tb-pdf}cardinalR additional functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_noisedims & Generates additional noise dimensions.\\
gen\_bkgnoise & Adds background noise.\\
randomize\_rows & Randomizes the rows.\\
relocate\_clusters & Relocates the clusters.\\
gen\_nproduct & Generates a vector of positive integers whose product is approximately equal to a target value.\\
\addlinespace
gen\_nsum & Generates a vector of positive integers whose summation is approximately equal to a target value.\\
gen\_wavydims1 & Generates random noise dimensions with wavy pattern generated with theta.\\
gen\_wavydims2 & Generates random noise dimensions with wavy pattern generated with power functions.\\
gen\_wavydims3 & Generates random noise dimensions with wavy pattern generated with power and sine functions.\\
gen\_rotation & Generates rotations.\\
\addlinespace
normalize\_data & Normalizes data.\\
\bottomrule
\end{tabular}
\end{table}

\section{Application}\label{application}

\subsection{Assessing the performance of dimension reduction on different geometric structures in high-dimensions}\label{assessing-the-performance-of-dimension-reduction-on-different-geometric-structures-in-high-dimensions}

We illustrate the use of the package by generating five clusters of different shapes, positioned with equal inter-cluster distances in \(4\text{-}D\). Using this synthetic dataset, we will evaluate the performance of six popular dimension reduction techniques: Principal Component Analysis (PCA) \citep{jolliffe2011}, t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}.

\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/unnamed-chunk-13-1}
